Beyond Static Retrieval: Adaptive Method for Improving RAG Performance

### 2025-02-19 Paper/latex/paper_RAG.tex
last update

\abstract{Retrieval-Augmented Generation (RAG) improves the quality of language model responses by incorporating external knowledge. However, its performance heavily relies on choosing the right number of documents to retrieve (K). Using a fixed number of documents (K) is computationally efficient but fails to adapt to the complexity of different queries. On the other hand, dynamically adjusting K offers flexibility but comes with higher computational costs and inconsistency. To overcome these challenges, we introduce a hybrid approach to dynamic K selection. This method adapts K based on the specific characteristics of each query while keeping computational demands manageable. Our strategy enhances retrieval performance by minimizing irrelevant information, ensuring the retrieval of relevant data, and boosting overall system accuracy. Experiments show that this approach outperforms both fixed and fully dynamic methods, offering a promising way to optimize retrieval in RAG systems.Future research will focus on developing learning-based techniques to further improve the dynamic selection of K.}

@@

Large Language Models (LLMs)\cite{Radford2019} have emerged as state-of-the-art artificial intelligence systems that excel at understanding and generating human-like text,Built on deep learning architectures,particularly transformers \cite{vaswani2017attention}.these models are trained on massive datasets to capture linguistic patterns and semantic relationships showcasing impressive adaptability across diverse tasks such as  machine translation\cite{Gu2018SearchEngine}, text generation\cite{liang2024controllabletextgenerationlarge} and question answering\cite{Zhang2023}.Notable examples of LLMs include OpenAI GPT series\cite{brown2020language} , Google BERT\cite{devlin2018bert} and T5 \cite{raffel2020exploring} which have set new benchmarks in NLP tasks .\\
Despite their impressive capabilities,LLMs suffer from hallucination\cite{ji2023survey} where they confidently generate factually incorrect or entirely fabricated information. This poses challenges in applications requiring high accuracy like knowledge-intensive tasks\cite{kandpal2023large} , Additionally significant limitation of LLMs is their static nature, as they are trained on fixed datasets\cite{bender2021dangers} and cannot dynamically update their knowledge which makes them prone to generating outdated or inconsistent responses, especially for time-sensitive or evolving data\cite{Mousavi2025}.

Researchers have explored various approaches To address this limitation, including transfer learning\cite{zhuang2020comprehensive}a technique that improves pre-trained models performance in new tasks by transferring knowledge from related tasks.and fine-tuning \cite{Howard2018} which adapt pre-trained models to specific tasks or domains. While these methods improve performance on targeted tasks,they are computationally expensive and impractical for real-time knowledge updates. A promising alternative is Retrieval-Augmented Generation (RAG) \cite{lewis2020retrieval}.which integrates real-time retrieval from external knowledge sources with generative models.

RAG enables LLMs to access up-to-date information without retraining, making it particularly useful for applications like open-domain question answering and fact-checking.
Retrieval-Augmented Generation (RAG) faces several limitations, The performance of RAG heavily depends on the quality of the retriever \cite{karpukhin2020dpr}, as irrelevant or noisy documents can degrade the generator output. Additionally, retrieving and processing a large number of documents increases computational overhead , making RAG less efficient for real-time applications. A critical challenge is determining the optimal number of retrieved passages (K) to feed into the generator\cite{lewis2020retrieval}. often resulting in either insufficient or excessive information being retrieved, reduce coherence and increase computational overhead. This trade-off makes K a key factor in balancing performance and efficiency.

Current approaches, such as static K selection \cite{lewis2020retrieval} involves retrieving a fixed number of top results regardless of the query characteristics ,it cannot adapt to the complexity of individual queries , dynamic K adjustment attempt to address this trade-off but often introduce complexity or latency ,Hybrid approaches like those proposed yuan et al . \cite{yuan2024hybrid},that integrates static retrieval with dynamic retrieval to improve accuracy. Furthermore, the "Blended RAG"\cite{sawarkar2024blendedragimprovingrag} approach \cite{blended_rag2024} enhances retrieval accuracy by combining semantic search techniques with hybrid query strategies. However, the approach faces significant limitations, including high computational overhead due to the need for multiple retrieval methods and dependency on high-quality metadata for optimal performance. These limitations highlight the need for a more robust approach to K selection, capable of balancing efficiency and accuracy across varying query types.

@@

Our contributions in this paper are the following :
we review existing approaches to Retrieval-Augmented Generation (RAG), with a specific focus on static , dynamic and Hybrid selection strategies. We critically analyze the strengths and limitations of these methods, identifying key gaps in their design and performance. These gaps motivate the need for an improved approach that addresses the challenges of both static and dynamic in retrieval-augmented systems. \\
Overview of Retrieval-Augmented Generation (RAG) in This section We provide a detailed explanation of the RAG framework, including its components and how it operates. Key concepts, such as retrievers, generators are discussed to establish the foundational context.\\
Proposed Solution: This section introduces our novel hybrid dynamic  selection algorithm. We describe its methodology, explain its design, and discuss how it overcomes the limitations of existing static and dynamic approaches. The benefits of this approach, particularly in improving retrieval accuracy and computational efficiency, are emphasized.\\
Results :in This section We present experimental results to demonstrate the effectiveness of our hybrid dynamic  selection algorithm. Comparisons with baseline method highlight its superiority in terms of retrieval performance and adaptability to diverse query complexities.

@@

In traditional Retrieval-Augmented Generation (RAG) systems, static K-selection is a widely used method where a fixed number of top-k documents are retrieved based on their relevance scores. Sparse retrieval methods where the "keys" represent documents to be retrieved, and the "values" are the same documents in this context such as query likelihood models\cite{Lafferty2001} ,TF-IDF\cite{Robertson1997} and BM25\cite{Zaragoza2009} often rely on this static approach to identify and rank documents. For instance, if k  =5, the system retrieves the top 5 documents deemed most relevant to a query, regardless of the query's complexity or the variability in relevance distribution among documents.
A major drawback of this approach is its rigidity. Since the number of retrieved documents (k) is fixed, it fails to adapt to the varying complexity of queries. For simple queries, retrieving too many documents may introduce noise, while for complex queries, retrieving too few may result in incomplete or insufficient information. This one-size-fits-all strategy can lead to suboptimal performance across diverse datasets or tasks.\\
\textbf{Dynamic top-k Selection in Retrieval}

Dynamic k-selection addresses the limitations of static k by varying the number of retrieved documents based on the specific query or context. These approaches typically involve heuristics, machine learning models, or adaptive algorithms to determine the optimal K for each query such as Dynamic Query Cutoff (DQC)\cite{Culpepper2016} This approach uses machine learning models trained on query-document features to predict the optimal k for retrieval , A reinforcement learning-based approach\cite{Zhou2021} where the agent learns to adjust k by maximizing downstream task performance (e.g accurate response generation in RAG) Another example is the Dynamic Selection of k Nearest Neighbors \cite{DynamicKNN2012} where the selection process often depends on the distribution of neighbors, their distances to the query point, or their relevance scores.
However, these methods come with limitations.They often introduce computational overhead due to the need for dynamic evaluation, which can increase latency in large-scale systems. Additionally, their performance heavily depends on the quality of the underlying models or distance metrics, making them sensitive to noisy data or poorly tuned hyperparameters. As a result, achieving generalization across diverse datasets or domains remains a significant challenge. \\
\textbf{Hybrid Approaches}

@@

\begin{figure}[h]
	\centering
	\includegraphics[width=0.9\textwidth]{rag.PNG}
	\caption{RAG system component }\label{rag.PNG}
\end{figure}
As illustrated in figure \ref{rag.PNG}, the RAG system is built around two main components:the retriever and the generator. The retriever?s role is to find relevant information from a pre-built data store, while the generator utilizes this retrieved data to produce coherent and contextually appropriate content. The overall RAG process operates as follows:
\subsection{Retrieval in RAG Systems}\label{subsec2}

@@

Depending on the similarity function used, retrieval methods can be grouped into categories such as sparse retrieval, dense retrieval, and others. For the widely used sparse and dense retrieval approaches, the process typically involves two main steps:  \\
1. Encoding each object into a specific representation.\\
2. Building an index to organize the data for efficient searching .

@@

\subsection{Generator  in RAG Systems}\label{subsec2}
In Retrieval-Augmented Generation (RAG) systems, the generator is a key component responsible for crafting the final output. It works by combining the retrieved information with the users input query to produce a well-structured and contextually relevant response .

@@

In various generative tasks, different models are selected based on the specific requirements: Models like BART\cite{lewis2019bart} and GPT\cite{brown2020language}are commonly used for tasks such as text summarization and translation ,Vision-Language Models (VLMs) \cite{radford2021learningtransferablevisualmodels} are employed to generate textual descriptions from images , for ext-to-Code Generation: Models like OpenAIs Codex \cite{chen2021evaluatinglargelanguagemodels} are designed to translate natural language prompts into executable code.

\section{proposed Slution}\label{sec3}
In traditional candidate selection systems, the reliance on static or fixed top-K retrieval lead to suboptimal retrieval performance, especially in scenarios where the relevance distribution of candidates is highly variable. To address these limitations, we propose Dynamic Candidate Selection, a novel algorithm that adaptively refines the candidate selection process by leveraging component-level embeddings and a Mixture of Logits (MoL) scoring mechanism.

The \textbf{Mixture of Logits (MoL)}\cite{zhai2023revisiting} methode provides a flexible and adaptive mechanism for computing similarity scores between queries and items. Given a query \( q \) and an item \( x \), MoL assumes that both are mapped to \( P \) groups of low-rank embeddings, denoted as \( f_p(q) \) and \( g_p(x) \), respectively. These embeddings are parameterized by neural networks that extract features from the query and item. The similarity between \( q \) and \( x \) is computed as a weighted sum of the inner products of these embeddings, where the weights \( \pi_p(q, x) \)\cite{zhai2023revisiting} are adaptive gating parameters constrained to sum to 1:
\[
\text{Similarity}(q, x) = \sum_{p=1}^P \pi_p(q, x) \cdot \langle f_p(q), g_p(x) \rangle.
\]
This formulation allows MoL to capture complex relationships between queries and items, making it particularly suitable for dynamic candidate selection tasks.

Inspired by the approach in\cite{ding2024efficient}, we introduce an adaptive threshold mechanism (\textbf{Dynamic Candidate Selection}), the MoL framework is employed to refine the candidate retrieval process. Specifically:
\begin{enumerate}
	\item \textbf{Component-Level Embeddings:}  
	 generate component-level embeddings for all items in \( X \). These embeddings allow efficient similarity computations during retrieval.
	\begin{equation*}
		X_p \leftarrow \{ g_p(x) \mid x \in X \}
	\end{equation*}

@@

	\item \textbf{Refinement and Top-K Selection:}  
	Additional relevant candidates are retrieved using the adaptive threshold \( T_{\text{adaptive}} \), expanding the candidate set \( G' \). The algorithm then selects the most relevant top-k results by sorting \( G' \) based on MoL scores.
	\begin{equation*}
		G' \leftarrow G \cup \{ x \mid s_p \geq T_{\text{adaptive}} \}
	\end{equation*}
	Finally, we sort candidates by MoL scores and extract the exact top-k:
	\begin{equation*}
		G_{\text{final}} = \text{Top-}k(G', \varphi(q, x))
	\end{equation*}
\end{enumerate}
the flowchart provided in Figure~\ref{fig2} describes the step-by-step process of the proposed algorithm. It visualizes the key stages of the solution
\begin{figure}[H]
	\centering
	\includegraphics[width=0.9\textwidth]{Flowchart_Steps.PNG}
	\caption{Algorithm workflow. }\label{fig2}
\end{figure}

As shown in Algorithm~\ref{Algo1}, the retrieval process dynamically adjusts the threshold based on MoL scores.
\begin{algorithm}
	\caption{Hybrid Exact Top-k with Threshold-Based k Selection}\label{ALgo1}
	\begin{algorithmic}[1]
		\State \textbf{Input:}
		\begin{itemize}
			\item Query $q$
			\item Set of items $X$
			\item Component-level embeddings: $f_p(q)$, $g_p(x)$ for $p \in P$, $x \in X$
			\item Initial threshold $T_{\text{init}}$
		\end{itemize}

		\State \textbf{Output:}
		\begin{itemize}
			\item Exact top $k$ items based on dynamic threshold selection, $G_{\text{final}}$
		\end{itemize}

		\State \textbf{1. Initialize:}
		\begin{itemize}
			\item Set $G \gets \emptyset$ \Comment{Initial candidate set}
		\end{itemize}

		\State \textbf{2. Generate Component-Level Embeddings:}
		\For{each component $p \in P$}
		\State $X_p \gets \{g_p(x) \mid x \in X\}$ \Comment{Precompute embeddings}
		\EndFor

		\State \textbf{3. Initial Candidate Retrieval:}
		\For{each component $p \in P$}
		\State Compute dot product scores: 
		\[
		S_p = \{ \langle f_p(q), g_p(x) \rangle : x \in X_p \}
		\]
		\State Retrieve items with scores $S_p \geq T_{\text{init}}$
		\State Add these items to $G$
		\EndFor

		\State \textbf{4. Adjust k Dynamically:}
		\For{each $x \in G$}
		\State Compute MoL scores $s = \phi(q, x)$ for each $x \in G$ using:
		\[
		\phi(q, x) = \sum_{p=1}^{P} \pi_p(q, x) \cdot \langle f_p(q), g_p(x) \rangle
		\]
		\State Set $T_{\text{adaptive}} = \min \{s : s \in G\}$
		\EndFor

		\State \textbf{5. Refine Candidate Set with Adaptive k:}
		\For{each component $p \in P$}
		\State Retrieve items from $X_p$ with scores $S_p \geq T_{\text{adaptive}}$
		\State Add these items to $G'$
		\EndFor

		\State \textbf{6. Select Exact Top-k Items:}
		\State Compute MoL scores for all items in $G'$
		\State Sort $G'$ by MoL scores in descending order
		\State Select the top $k$ items from $G'$ where $k$ is the number of items in $G'$ exceeding $T_{\text{adaptive}}$

		\State \textbf{7. Return:} $G_{\text{final}}$ \Comment{Top $k$ items from $G'$}
	\end{algorithmic}
\end{algorithm}
\section{Experimental Results}\label{sec3}
We evaluate the performance of our hybrid retrieval algorithm and the Mixture-of-Logits (MoL) approach on the next-item prediction task in recommendation systems\cite{Zhu_2018}\cite{kang2018selfattentivesequentialrecommendation}. We conducted experiments using two MovieLens datasets: ML-100K and ML-1M \cite{Harper2015}  which are widely used benchmarks for evaluating sequential recommendation models.\\
\textbf{MovieLens-100K} \\
100,000 ratings (1-5 scale) from 943 users on 1,682 movies\\
Each user has rated at least 20 movies \\
Includes user demographic information (age, gender, occupation, zip code)\\
\textbf{MovieLens-1M} \\
1,000,209 ratings from 6,040 users on approximately 3,900 movies \\
Data collected from users who joined MovieLens in 2000 \\
Represents a larger-scale recommendation scenario \\
\subsection{ Setup }
For both datasets, We utilize the SASRec architecture as our sequential user encoder, a model renowned for achieving state-of-the-art performance in next-item prediction tasks. This architecture processes the user's historical interaction sequence, generating embeddings that encapsulate the user's preferences at each time step. These embeddings serve as the foundation for predicting the next item in the sequence.\\

>>>>

\abstract{Retrieval-Augmented Generation (RAG) improves the quality of language model responses by incorporating external knowledge. However, its performance heavily relies on choosing the right number of documents to retrieve (K).therefore,despite it being computationally efficient,  Using a fixed number of documents (K) fails to adapt to the complexity of different queries.On the other hand, dynamically adjusting K offers flexibility but comes with higher computational costs and inconsistency. To overcome these challenges, we introduce a hybrid approach to dynamic K selection. This method adapts K based on the specific characteristics of each query while keeping computational demands manageable. Consequently, our  strategy enhances retrieval performance by minimizing irrelevant information, ensuring the retrieval of relevant data, and boosting overall system accuracy. Experiments show that this approach outperforms both fixed and fully dynamic methods, offering a promising way to optimize retrieval in RAG systems.Future research will focus on developing learning-based techniques to further improve the dynamic selection of K.}

@@

Large Language Models (LLMs)\cite{Radford2019} have emerged as state-of-the-art artificial intelligence systems that excel at understanding and generating human-like text,Built on deep learning architectures,particularly transformers \cite{vaswani2017attention}. These models are trained on massive datasets to capture linguistic patterns and semantic relationships, thereby showcasing impressive adaptability across diverse tasks, such as machine translation\cite{Gu2018SearchEngine}, text generation\cite{liang2024controllabletextgenerationlarge} and question answering\cite{Zhang2023}.Notable examples of LLMs include OpenAI GPT series\cite{brown2020language} , Google BERT\cite{devlin2018bert} and T5 \cite{raffel2020exploring} which have set new benchmarks in NLP tasks .\\
Despite their impressive capabilities,LLMs suffer from hallucination\cite{ji2023survey} where they confidently generate factually incorrect or entirely fabricated information.Naturally,This poses significant challenges in applications requiring high accuracy like knowledge-intensive tasks\cite{kandpal2023large}.Another limitation of LLMs is their static nature, as they are trained on fixed datasets\cite{bender2021dangers} and cannot dynamically update their knowledge . As a result, they are prone to generate outdated or inconsistent responses, especially for time-sensitive or evolving data\cite{Mousavi2025}.

Researchers have explored various approaches to address this limitation, including transfer learning\cite{zhuang2020comprehensive}a technique that improves pre-trained models performance in new tasks by transferring knowledge from related tasks.Another technique is fine-tuning \cite{Howard2018} which adapt pre-trained models to specific tasks or domains. While these methods improve performance on targeted tasks,they are computationally expensive and impractical for real-time knowledge updates. A promising alternative is Retrieval-Augmented Generation (RAG) \cite{lewis2020retrieval}.which integrates real-time retrieval from external knowledge sources with generative models.

RAG enables LLMs to access up-to-date information without retraining, making it particularly useful for applications like open-domain question answering and fact-checking.However,(RAG) faces several limitations. For instance,The performance of RAG heavily depends on the quality of the retriever \cite{karpukhin2020dpr}, as irrelevant or noisy documents can degrade the generator output. Additionally, retrieving and processing a large number of documents increases computational overhead , making RAG less efficient for real-time applications. A critical challenge is determining the optimal number of retrieved passages (K) to feed into the generator\cite{lewis2020retrieval}. often resulting in either insufficient or excessive information being retrieved, reduce coherence and increase computational overhead. This trade-off makes K a key factor in balancing performance and efficiency.

Current approaches, such as static K selection \cite{lewis2020retrieval} involves retrieving a fixed number of top results regardless of the query characteristics ,it cannot adapt to the complexity of individual queries , dynamic K adjustment attempt to address this trade-off but often introduce complexity or latency ,Hybrid approaches like those proposed yuan et al . \cite{yuan2024hybrid},that integrates static retrieval with dynamic retrieval to improve accuracy. Furthermore, the "Blended RAG"\cite{sawarkar2024blendedragimprovingrag} approach enhances retrieval accuracy by combining semantic search techniques with hybrid query strategies. However, the approach faces significant limitations, including high computational overhead due to the need for multiple retrieval methods and dependency on high-quality metadata for optimal performance. These limitations highlight the need for a more robust approach to K selection, capable of balancing efficiency and accuracy across varying query types.

@@

This paper contributes to the existing literature by reviewing the various approaches to Retrieval-Augmented Generation (RAG), with a specific focus on static, dynamic, and Hybrid selection strategies. Additionally, it provides a critical analysis of the strengths and limitations of these methods, as it identifies key gaps in their design and performance. On account of this, there arises a pressing need for an improved approach capable of addressing the challenges of both static and dynamic in retrieval-augmented systems. \\
The first section, Overview of Retrieval-Augmented Generation (RAG), aims to provide an a detailed explanation of the RAG framework, including its components and how it operates. Key concepts, such as retrievers, generators are discussed to establish the foundational context.\\
The following section,introduces the novel hybrid dynamic selection algorithm. Additionally, it provides detailed description of its methodology, explains its design, and discusses the manner in which it overcomes the limitations of existing static and dynamic approaches. Particular emphasis is placed on the benefits of this approach, particularly in improving retrieval accuracy and computational efficiency.\\
The final section presents experimental results that demonstrate the effectiveness of the hybrid dynamic selection algorithm.highlight its superiority in terms of retrieval performance and adaptability to diverse query complexities

@@

In traditional Retrieval-Augmented Generation (RAG) systems, static K-selection is a widely used method where a fixed number of top-k documents are retrieved based on their relevance scores. Sparse retrieval\cite{bai2020spartermlearningtermbasedsparse} methodsw here documents and queries are represented as high-dimensional, sparse vectors In these method , the "keys" represent documents to be retrieved, and the "values" are he importance or weight of those terms in the document such as query likelihood models\cite{Lafferty2001} ,TF-IDF\cite{Robertson1997} and BM25\cite{Zaragoza2009} often rely on this static approach to identify and rank documents. For instance, if k  =5, the system retrieves the top 5 documents deemed most relevant to a query, regardless of the query's complexity or the variability in relevance distribution among documents.
A major drawback of this approach is its rigidity. Since the number of retrieved documents (k) is fixed, it fails to adapt to the varying complexity of queries. For simple queries, retrieving many documents may introduce noise, while for complex queries, retrieving few documents may result in incomplete or insufficient information. This one-size-fits-all strategy can lead to suboptimal performance across diverse datasets or tasks.\\
\textbf{Dynamic top-k Selection in Retrieval}

Dynamic k-selection addresses the limitations of static k by varying the number of retrieved documents based on the specific query or context. These approaches typically involve heuristics, machine learning models, or adaptive algorithms to determine the optimal K for each query such as Dynamic Query Cutoff (DQC)\cite{Culpepper2016} This approach uses machine learning models trained on query-document features to predict the optimal k for retrieval , A reinforcement learning-based approach\cite{Zhou2021} where the agent learns to adjust k by maximizing downstream task performance (e.g accurate response generation in RAG) Another example is the Dynamic Selection of k Nearest Neighbors \cite{DynamicKNN2012} where the selection process often depends on the distribution of neighbors, their distances to the query point, or their relevance scores.
However, these methods come with limitations.They often introduce computational overhead due to the need for dynamic evaluation, which can increase latency in large-scale systems. Additionally, their performance heavily depends on the quality of the underlying models or distance metrics, making them sensitive to noisy data or poorly tuned hyperparameters. As a result, achieving generalization across diverse datasets or domains remains a significant challenge. \\
\textbf{Hybrid Approaches For Top-K selection in Retrieval}

@@

\begin{figure}[h]
	\centering
	\includegraphics[width=0.9\textwidth]{rag.PNG}
	\caption{RAG system component\cite{sharma2024rag}. }\label{rag.PNG}
\end{figure}
As illustrated in figure \ref{rag.PNG}, the RAG system is built around two main components:the retriever and the generator. The retriever's role is to find relevant information from a pre-built data store, while the generator utilizes this retrieved data to produce coherent and contextually appropriate content. The overall RAG process operates as follows :
\subsection{Retrieval in RAG Systems}\label{subsec2}

@@

Depending on the similarity function used, retrieval methods can be grouped into categories such as sparse retrieval\cite{bai2020spartermlearningtermbasedsparse}, dense retrieval\cite{karpukhin2020dpr}, and others. For the widely used sparse and dense retrieval approaches, the process typically involves two main steps:  \\
1. Encoding each object into a specific representation.\\
2. Building an index to organize the data for efficient searching .

@@

\subsection{Generator  in RAG Systems}\label{subsec2}
In Retrieval-Augmented Generation (RAG) systems, the generator\cite{huang2024surveyretrievalaugmentedtextgeneration} is a key component responsible for crafting the final output. It works by combining the retrieved information with the users input query to produce a well-structured and contextually relevant response .

@@

In various generative tasks, different models are selected based on the specific requirements: Models like BART\cite{lewis2019bart} and GPT\cite{brown2020language}are commonly used for tasks such as text summarization and translation ,Vision-Language Models (VLMs) \cite{radford2021learningtransferablevisualmodels} are employed to generate textual descriptions from images , for text-to-Code Generation: Models like OpenAIs Codex \cite{chen2021evaluatinglargelanguagemodels} are designed to translate natural language prompts into executable code.

\section{Proposed Slution}\label{sec3}
In traditional candidate selection systems, the reliance on static or fixed top-K retrieval lead to suboptimal retrieval performance, especially in scenarios where the relevance distribution of candidates is highly variable. In this section, we introduce the Mixture of Logits (MoL) and describe its role in representing a learned similarity function. Table \ref{tab:Notations_table} summarizes the notations used in this paper and we propose Dynamic Candidate Selection, a novel algorithm that adaptively refines the candidate selection process by leveraging component-level embeddings and a Mixture of Logits (MoL) scoring mechanism.

\subsection{Mixture of Logits}

The Mixture of Logits (MoL)\cite{zhai2023revisiting} methode provides a flexible and adaptive mechanism for computing similarity scores between queries and items. Given a query \( q \) and an item \( x \), MoL assumes that both are mapped to \( P \) groups of low-rank embeddings, denoted as \( f_p(q) \) and \( g_p(x) \), respectively. These embeddings are parameterized by neural networks that extract features from the query and item. The similarity between \( q \) and \( x \) is computed as a weighted sum of the inner products of these embeddings, where the weights \( \pi_p(q, x) \)\cite{zhai2023revisiting} are adaptive gating parameters constrained to sum to 1:
\[
\text{Similarity}(q, x) = \sum_{p=1}^P \pi_p(q, x) \cdot \langle f_p(q), g_p(x) \rangle.
\]
This formulation allows MoL to capture complex relationships between queries and items, making it particularly suitable for dynamic candidate selection tasks.

\begin{table}[h!]
	\centering
	\normalsize 
	\resizebox{\textwidth}{!}{ 
	\begin{tabular}{|c|l|}
		\hline
		\textbf{Notation} & \textbf{Description} \\ \hline
		\( q \) (\( Q \), \( |Q| \)) & A single query ,(set of all queries, total number of queries). \\ \hline
		\( x \) (\( X \), \( |X| \)) & A single item ,(set of all items, total number of items). \\ \hline
		\( \phi(q, x) \) & The learned similarity function, Mixture-of-Logits (MoL), which computes the similarity between \( q \) and \( x \). \\ \hline
		\( P \) (\( P_q \), \( P_x \)) & Total number of low-rank embeddings (\( P = P_q \times P_x \)), where \( P_q \) and \( P_x \) are the number of embeddings for \( q \) and \( x \), respectively. \\ \hline
		\( \pi_p(q, x) \)  & Weight for the \( p \)-th (or \( p_q \)-th and \( p_x \)-th) embedding set for \( (q, x) \). \\ \hline
		\( f(q) \) (\( f_p(q) \)) & Learned embedding for the query (\( p \)-th component-level embedding for \( q \)). \\ \hline
		\( g(x) \) (\( g_p(x) \)) & Learned embedding for the item (\( p \)-th component-level embedding for \( x \)). \\ \hline
		\( \langle f(q), g(x) \rangle \) & Dot product similarity function: \( \langle f(q), g(x) \rangle = g(x)^T f(q) \). \\ \hline
	\end{tabular} }
	\caption{ Notations table}
	\label{tab:Notations_table}
\end{table}
\subsection{Retrieval Algorithm}
Inspired by the approach in\cite{ding2024efficient}, we introduce an adaptive threshold mechanism (\textbf{Dynamic Candidate Selection}), the MoL framework is employed to refine the candidate retrieval process. Specifically:
\begin{enumerate}
	\item \textbf{Component-Level Embeddings:}  
	 generate component-level embeddings for all items in \( X \). These embeddings allow efficient similarity computations during retrieval.
	\begin{equation*}
		X_p \leftarrow \{ g_p(x) \mid x \in X \}
	\end{equation*}

@@

	\item \textbf{Refinement and Top-K Selection:}  
	Additional relevant candidates are retrieved using the adaptive threshold \( T_{\text{adaptive}} \), expanding the candidate set \( G' \). The algorithm then selects the most relevant top-k results by sorting \( G' \) based on MoL scores.
	\begin{equation*}
		G' \leftarrow G \cup \{ x \mid s_p \geq T_{\text{adaptive}} \}
	\end{equation*}
	\item \textbf{ Select Exact Top-k Items:} Finally, we sort candidates by MoL scores and extract the exact top-k:
	\begin{equation*}
		G_{\text{final}} = \text{Top-}k(G', \varphi(q, x))
	\end{equation*}
\end{enumerate}
the flowchart provided in Figure~\ref{fig2} describes the step-by-step process of the proposed algorithm. It visualizes the key stages of the solution
\begin{figure}[H]
	\centering
	\includegraphics[width=0.9\textwidth]{Flowchart_Steps.PNG}
	\caption{Algorithm workflow. }\label{fig2}
\end{figure}

As shown in Algorithm\ref{ALgo1}, the retrieval process dynamically adjusts the threshold based on MoL scores.
\begin{algorithm}
\caption{Hybrid Exact Top-k with Threshold-Based k Selection}
\label{ALgo1}
	\begin{algorithmic}[1]
     \textbf{Input:}
	     	 Query $q$ , Set of items $X$ ,Component-level embeddings: $f_p(q)$, $g_p(x)$ for $p \in P$, $x \in X$ , Initial threshold $T_{\text{init}}$  \newline
             \textbf{Output:}
	         Exact top $k$ items based on dynamic threshold selection, $G_{\text{final}}$ \newline

         	\State Set $G \gets \emptyset$     \Comment{Initial candidate set}
		\For{each component $p \in P$}
		\State $X_p \gets \{g_p(x) \mid x \in X\}$ \Comment{Precompute embeddings}
		\EndFor

	    \textbf{1. Initial Candidate Retrieval:}
		\For{each component $p \in P$}
		\State Compute dot product scores: 
		\[
		S_p \gets  \{ \langle f_p(q), g_p(x) \rangle : x \in X_p \}
		\]
		\State Retrieve items with scores $S_p \geq T_{\text{init}}$
		\State Add these items to $G$
		\EndFor

		 \textbf{2. Adjust k Dynamically:}
		\For{each $x \in G$}
		\State Compute MoL scores $s \gets \phi(q, x)$ where :
		\[
		\phi(q, x) \gets \sum_{p=1}^{P} \pi_p(q, x) \cdot \langle f_p(q), g_p(x) \rangle
		\]
		\State Set $T_{\text{adaptive}} = \min \{s : s \in G\}$
		\EndFor

		\textbf{3. Refine Candidate Set with Adaptive k:}
	    \State	$G' \gets \emptyset$
		\For{each component $p \in P$}
		\State Retrieve items from $X_p$ with scores $S_p \geq T_{\text{adaptive}}$
		\State Add these items to $G'$
		\EndFor

		 \textbf{4. Select Exact Top-k Items:}
		\State Compute MoL scores for all items in $G'$
		\State Sort $G'$ by MoL scores in descending order
		\State Select the top $k$ items from $G'$ where $k$ is the number of items in $G'$ exceeding $T_{\text{adaptive}}$

		\State\textbf{ Return:} $G_{\text{final}}$ \Comment{Retrieve Top $k$ items from $G'$}
	\end{algorithmic}
\end{algorithm}
\section{Experimental Results}\label{sec3}
We evaluate the performance of our hybrid retrieval algorithm and the Mixture-of-Logits (MoL) approach on the next-item prediction task in recommendation systems\cite{Zhu_2018}\cite{kang2018selfattentivesequentialrecommendation}. We conducted experiments using two MovieLens datasets: ML-100K and ML-1M \cite{Harper2015}  which are widely used benchmarks for evaluating sequential recommendation models.\\
\textbf{MovieLens-100K} \\
100,000 ratings (1-5 scale) from 943 users on 1,682 movies\\
Each user has rated at least 20 movies \\
Includes user demographic information (age, gender, occupation, zip code)\\
\textbf{MovieLens-1M} \\
1,000,209 ratings from 6,040 users on approximately 3,900 movies \\
Data collected from users who joined MovieLens in 2000 \\
Represents a larger-scale recommendation scenario \\
\subsection{ Setup }
For both datasets, We utilize the SASRec architecture as our sequential user encoder, a model renowned for achieving state-of-the-art performance in next-item prediction tasks. This architecture processes the user's historical interaction sequence, generating embeddings that encapsulate the user's preferences at each time step. These embeddings serve as the foundation for predicting the next item in the sequence.

The query q represents the user's state at a specific time step, derived from their interaction history. In the MoL (Mixture of Logits) framework, q is transformed into Pq
embeddings through a multi-layer perceptron (MLP). While our hybrid algorithm does not explicitly employ an MLP for this transformation
\subsection{Hyperparameter settings.}
For fair comparison, we maintained consistent architectural choices and training conditions across all experiments , We conducted an extensive hyperparameter analysis comparing both approaches (Hybrid+SAS and MoL+SAS) across different architectural configurations.All experiments were pmplemented in TensorFlow and trained on a Google Colab environment with a T4 GPU. We use the Adam optimizer with a learning rate of 0.001. For the hybrid algorithm, we initialize the threshold (Tinit) to 0.3 and adaptively adjust it during training. .We discuss detailed
hyperparameter settings Table\ref{tab:100k_movies},table\ref{tab:1m_movies}
\begin{table}[h]
 \centering
\normalsize 
\resizebox{\textwidth}{!}{ 
	\begin{tabular}{lccccccccc}
			\toprule
			\textbf{Model} & \textbf{Max Sequence Length} & \textbf{Embedding Dimension} & \textbf{Number of Heads} & \textbf{Feedforward Dimension} & \textbf{Batch Size} & \textbf{Epochs} & \textbf{Val Loss} & \textbf{Val Accuracy} \\ 
			\midrule
			Hybrid+SAS & 50  & 128  & 2 & 128  & 128 & 12 & 5.3036 & 0.1584  \\
			MoL+SAS    & 50  & 128  & 2 & 128  & 128 & 12 & 5.3152 & 0.1582  \\
			Hybrid+SAS & 128 & 256  & 4 & 256  & 128 & 10 & 3.6364 & 0.4301  \\
			MoL+SAS    & 128 & 256  & 4 & 256  & 128 & 10 & 3.6374 & 0.4299 \\
			Hybrid+SAS & 512 & 512  & 4 & 512  & 128 & 10 & 1.2808 & 0.8120 \\
			MoL+SAS    & 512 & 512  & 4 & 512  & 128 & 10 & 1.3050 & 0.8016 \\
			\bottomrule
		\end{tabular}
	}
	\caption{Results on 100kMovies dataset}
	\label{tab:100k_movies}
\end{table}
